---
title: "Adding components to the model"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Rgadget)
gd <- gadget.variant.dir('simple_model')
```

## Gadget variables 

It can be tedious to change all the input files whenever you want to try a different value for a certain input parameter. For example if you want to specify natural mortality as some thing else than 0 you can always do:
```{r}
gadgetstock('simple_stock',gd) %>% 
  gadget_update('naturalmortality',0.2)
```
And if you want to try many version of natural mortality you can obviously repeat this process for different values of M. There is however a more convenient approach to do this, namely Gadget switches or parameters. A Gadget parameter can be inserted instead of any number in the input file (apart from likelihood data), and are marked with the `#` character. So for the case of natural mortality we can now update the stock file to include a Gadget parameter by:

```{r}
gadgetstock('simple_stock',gd) %>% 
  gadget_update('naturalmortality',"#M")
```

And now we see that the value for natural mortality has been changed from 0 to `#M`. Now write this to file and rerun run Gadget.
```{r echo=FALSE}
gadgetstock('simple_stock',gd) %>% 
  gadget_update('naturalmortality','#M') %>% 
  write.gadget.file(gd)

callGadget(s=1,main='main',ignore.stderr = FALSE,log = 'simple_log')
```

After running this, assuming we did everything correctly, we should see a new file in the directory called `params.out`:
```{r echo=FALSE}
read_lines('simple_model/params.out') %>% 
  paste(collapse = '\n') %>% 
  cat()
```

This file is generated by Gadget based on the switches it encounters when reading in the input files. It lists all parameters, their value as used in the simulation, upper and lower bounds and whether the parameter should optimised (more on that later). You then simply just change the value for M and re-run Gadget by supplying the parameter file as input. This can either be done manually or using the `init_guess` function:
```{r}
read.gadget.parameters('simple_model/params.out') %>% 
  init_guess('M',value = 0.2) %>% 
  write.gadget.parameters(file = 'simple_model/params.in')

callGadget(s=1,i='params.in')
```

Note that when supplying a parameter input you use the `i` switch in `callGadget`. 

Gadget parameter have a few more tricks up their sleave. Notably you can apply functions or formulas to the parameter. For instance if you want multiply the value of `M` by 0.1 you can write that as:
```
(* 0.1 #M)
```
Similarly if you need to log transform your variable you can:
```
(log #M)
```
and these can be chained together:
```
(* 0.1 (log #M))
```
allowing you to apply fairly elaborate functions to the parameter. Rgadget further provides mechanisms to work with Gadget formulae, both allowing you to evaluate the formula using `parse.gadget.formula` and transform a R statement with `to.gadget.formulae`:

```{r}
## R expression to Gadget formula
to.gadget.formulae(quote((1+exp(a*b))))

## Gadget formula to a R expression
parse.gadget.formulae('(* #a #b)')

## evaluate expression
parse.gadget.formulae('(* #a #b)') %>% 
  eval(list(a=7,b=6))
```

#### Excercise
* Change the initial conditions in such a way you can control the initial number of fish using the parameter file. 

## Fleet operations
Fleet operations are defined as a separate model component. 

```{r}
igfs.landings <- 
  structure(data.frame(year=schedule$year,step=2,area=1,number=1),
            area_group = list(`1` = 1)) 


gadgetfleet('Modelfiles/fleet',gd,missingOkay = TRUE) %>% 
    gadget_update('linearfleet',
                  name = 'igfs',
                  suitability = 
                    list(simple_stock=list(type='function',suit_func = 'constant',value=0.2),
                  data = igfs.landings) %>% 
  write.gadget.file(gd)
```

