---
title: "Defining multi area models"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE,message = FALSE)
```
# Defining multi area models

## Alternate structures: multiple areas

Let's return to the initial set-up of the ling model and modify it such that
the stock live on two areas. The first step to do this is to modify the area
file and stock files to include a second area.

```{r area2, message=FALSE}

library(Rgadget)
library(tidyverse)

#assuming you are beginning in a fresh session; this information should be in an 
#initialization script

base_dir <- 'ling_model'
vers <- c('01-base')
gd <- gadget.variant.dir(sprintf(paste0("%s/",vers),base_dir))

ling.imm <- gadgetstock('lingimm',gd)
ling.mat <- gadgetstock('lingmat',gd)
#area_file <- read.gadget.file(gd, 'Modelfiles/area', file_type = 'area')

## Write out areafile and update mainfile with areafile location

schedule <- 
  expand.grid(year = 1982:2018, step = 1:4, area = 1:2) %>% 
  arrange(year, step, area)

area_file <- 
  gadgetfile('Modelfiles/area',
             file_type = 'area',
             components = list(list(areas = c(1,2),
                                    size = c(1,1),
                                    temperature= schedule %>% mutate(temperature = 5)))) 
area_file %>% 
  write.gadget.file(gd)

ling.imm <-
  ling.imm %>% 
  gadget_update('stock',
                livesonareas = area_file[[1]]$areas)  
ling.mat <- 
  ling.mat %>% 
  gadget_update('stock',
                livesonareas = area_file[[1]]$areas)  

ling.imm
```

In theory, the only details left that are necessary to set up a two-area model
is to provide initial values and recruitment, as these are designated both by
area and age combinations:

```{r}

#taken from data analyses
init.sigma.i <- rep(2, length(ling.imm[[1]]$minage:ling.imm[[1]]$maxage))
init.sigma.m <- rep(2, length(ling.mat[[1]]$minage:ling.mat[[1]]$maxage))

ling.imm <-
  ling.imm %>%
  gadget_update('initialconditions',
                normalparam = 
                  expand.grid(age = ling.imm[[1]]$minage:ling.imm[[1]]$maxage, area = area_file[[1]]$areas) %>% 
                  as_tibble() %>% 
                  mutate( 
                    age.factor = parse(text=sprintf('exp(-1*(lingimm.M+ling.init.F)*%1$s)*lingimm.init.%1$s',age)) %>% 
                      map(to.gadget.formulae) %>% 
                      unlist(),   
                    area.factor = '#lingimm.init.scalar',
                    mean = von_b_formula(age,linf='ling.Linf',k='ling.k',recl='ling.recl'),
                    stddev = rep(init.sigma.i,2),
                    alpha = '#ling.walpha',
                    beta = '#ling.wbeta'
                    )
                ) %>%
  gadget_update('doesrenew',
                normalparam = 
                  expand.grid(year = schedule$year,area = area_file[[1]]$areas) %>%
                  distinct() %>% 
                  as_tibble() %>% 
                  mutate(step = 1,
                         age = ling.imm[[1]]$minage,
                         number = parse(text=sprintf('ling.rec.scalar*ling.rec.%s',year)) %>% 
                           map(to.gadget.formulae) %>% 
                           unlist(),
                         mean = von_b_formula(age,linf='ling.Linf',k='ling.k',recl='ling.recl'),
                         stddev = '#ling.rec.sd',
                         alpha = '#ling.walpha',
                         beta = '#ling.wbeta'
                         )
                ) 

ling.mat <-
  ling.mat %>% 
  gadget_update('initialconditions',
                normalparam = expand.grid(age = ling.mat[[1]]$minage:ling.mat[[1]]$maxage, area = area_file[[1]]$areas) %>% 
                  as_tibble() %>% 
                  mutate(
                    age.factor = parse(text=sprintf('exp(-1*(lingmat.M+ling.init.F)*%1$s)*lingmat.init.%1$s',age)) %>% 
                      map(to.gadget.formulae) %>% 
                      unlist(),
                    area.factor = '#lingmat.init.scalar',
                    mean = von_b_formula(age,linf='ling.Linf',k='ling.k',recl='ling.recl'),
                    stddev = rep(init.sigma.m,2),
                    alpha = '#lingmat.walpha',
                    beta = '#lingmat.wbeta'
                    )
                )  
## write to file
ling.imm %>% 
  write.gadget.file(gd)

ling.mat %>% 
  write.gadget.file(gd)
```

The only changes added to the above model were to add a second area with the
same exact values in initial conditions and recruitment as found in the
previous model with only a single area. The second area is therefore a
simulated mirror of the first, but with no fishing, as the fleets only 
exist on area 1. Structurally, this is the minimum possible information 
needed to include a second area.

```{r cache=TRUE,eval=FALSE}
gadget_evaluate(gd,log = 'log_area2')

```

### Migration

A model as shown in the previous section may be useful for theoretical
simulations, otherwise it is not particularly interesting, and has no 
effect on what goes on in area 1, including the model fitting procedures 
(as data are specified by area). A slightly more interesting model could
be a single population in two areas that can interact between areas 
via migration, which determines how much the individuals move among areas.

Migration is specified in the stock file under the migration section, 
and is therefore updated first by indicating that migration exists by 
turning the `doesmigrate` option from 0 to 1, and providing a migration 
matrix or migration ratios (two forms of the same information). In Rgadget,
this can be done by creating a migration matrix file and using 
`gadget_update` to modify the migration component of the stock file to 
refer to this file.

A migration matrix is a square n x n matrix for the n areas that the stock
is defined on with proportions moving *from* each area given in columns
*to* the area given in rows. Migration should not change the total number 
of fish in the model, so each column in the migration matrix should 
sum to 1. Therefore in the following example, at the start of migration, 
60% of individuals in area 1 remain there, 40% move to area 2. Likewise, 
30% of individuals in area 2 move to area 1, whereas 70% remain there:

```{r}
matrix(c(0.6, 0.4, 0.3, 0.7), nrow = 2)
```

An alternate method for supplying these values is via migration ratios, 
which instead lists only the proportions of fish moving, and lets Gadget
calculate the full matrix from those. To represent the above matrix in 
this form, we would write:
```{r}
matrix(c(1,2,0.4,2,1,0.3), nrow = 2, byrow = T)
```
In this form, within each row, the first column shows the origin area, the second the 
destination area, and the third the proportion. Note that migrations of 0% are calculated 
automatically, so they should be omitted.

In this example we continue with the migration ratios format. Ratios can be replaced with 
switches that are specified in the parameters file, and the matrix can 
be placed into a Gadget file with the necessary name and keyword
`[migrationmatrix]` using `gadgetfile`:

```{r}
tmp.mat <- matrix(c(1,2,0.4,2,1,'#mig'), nrow = 2, byrow = T)

#produce a migration matrix

    gadgetfile('Modelfiles/migration',
                 components = list(migrationmatrix = list(
                                        name = 'test',
                                        as.list(tmp.mat[1,]),
                                        as.list(tmp.mat[2,])
                                        )
                                   ) 
    )
   
```

Each migration file can include a series of migrations, but note that the order of evaluation of
these affects the ratios that should be written. For example, suppose we have 1000 fish in 
each area, and we want to represent the situation that 30% of the fish in area 1 will move to area 2 in time 
step 1, and the same ratio will migrate to area 1 from 2 in time step 2. We would write
the following:

```{r}

tmp.mat.w <- matrix(c(1,2,0.3), nrow = 1, byrow = T)
tmp.mat.e <- matrix(c(2,1,0.3), nrow = 1, byrow = T)

mig_file <-
    gadgetfile('Modelfiles/migration',
                 components = list(migrationmatrix = list(
                                        name = 'west',
                                        as.list(tmp.mat.w[1,])
                                        ),
                                   migrationmatrix =list(
                                        name = 'east',
                                        as.list(tmp.mat.e[1,])
                                        )
                                   )
               )

mig_file    
```

This indicates that 300 fish will move from area 1 to 2 in step 1, and then subsequently 
0.3 x 1300 = 390 fish will move from area 2 to area 1. It does *not* mean that 300 fish
will move back, or that these fish are the same individuals that moved in step 1. For 
this pattern a different model structure (with a new stock) is needed, so
that those 300 individuals can be tracked separately. 

For our purposes, let's set the migration proportions to switches:
```{r}

tmp.mat.w <- matrix(c(1,2,'#ling.mig.w'), nrow = 1, byrow = T)
tmp.mat.e <- matrix(c(2,1,'#ling.mig.e'), nrow = 1, byrow = T)

mig_file <-
    gadgetfile('Modelfiles/migration',
                 components = list(migrationmatrix = list(
                                        name = 'west',
                                        as.list(tmp.mat.w[1,])
                                        ),
                                   migrationmatrix =list(
                                        name = 'east',
                                        as.list(tmp.mat.e[1,])
                                        )
                                   )
               )

mig_file %>% 
      write.gadget.file(gd)

mig_file    
```

Although in the above example we have time steps for migration in mind, we have not yet
designated them in a Gadget file. To do this, the next step is to create a migration timing file:
```{r}
    gadgetfile('Modelfiles/migrationtime',
               file_type = 'data',
               components = list(bind_rows(
                 tibble(year = schedule$year %>% unique,
                            step = 2,
                            defineratios='east'),
                 tibble(year = schedule$year %>% unique,
                             step = 1,
                             defineratios='west')
                             ))) %>% 
    write.gadget.file(gd)

```

Finally, we modify the stock files for the stocks that should use this migration matrix and time files: 

```{r}

ling.imm <-
  ling.imm %>% 
    gadget_update('doesmigrate',
                  yearstepfile = 'Modelfiles/migrationtime',
                  defineratios = 'Modelfiles/migration')
ling.mat <-
  ling.mat%>% 
    gadget_update('doesmigrate',
                  yearstepfile = 'Modelfiles/migrationtime',
                  defineratios = 'Modelfiles/migration')

## write to file
ling.imm %>% 
  write.gadget.file(gd)

ling.mat %>% 
  write.gadget.file(gd)

ling.mat
```

```{r cache=TRUE,eval=FALSE}
gadget_evaluate(gd,log = 'log_area2')
```


### Using data from multiple areas

It is also possible to fit a Gadget model using area-specific data. 
If data are available from both areas, then there must be fleets operating
in both areas, so the first step is to modify the fleets. This is easily
done in Rgadget by including catch data from both areas and including 
area 2 in the attributes:

```{r fleets2}
lln.landings <- structure(
  bind_rows(read_csv('data_provided/lln.csv'),
            read_csv('data_provided/lln.csv') %>% 
              mutate(area=2,
                     total_weight = round(total_weight*rlnorm(n(), sdlog = 0.5))) %>%
                        filter(total_weight > 0)),
            area_group = list(`1` = 1, `2` = 2))
bmt.landings <- structure(
  bind_rows(read_csv('data_provided/bmt.csv'),
            read_csv('data_provided/bmt.csv') %>% 
              mutate(area=2,
                     total_weight = round(total_weight*rlnorm(n(), sdlog = 0.5))) %>%
                        filter(total_weight > 0)),
            area_group = list(`1` = 1, `2` = 2))
gil.landings <- structure(
  bind_rows(read_csv('data_provided/gil.csv'),
            read_csv('data_provided/gil.csv') %>% 
              mutate(area=2,
                     total_weight = round(total_weight*rlnorm(n(), sdlog = 0.5))) %>%
                        filter(total_weight > 0)),
            area_group = list(`1` = 1, `2` = 2))
for.landings <- structure(
  bind_rows(read_csv('data_provided/for.csv'),
            read_csv('data_provided/for.csv') %>% 
              mutate(area=2,
                     total_weight = round(total_weight*rlnorm(n(), sdlog = 0.5))) %>%
                        filter(total_weight > 0)),
            area_group = list(`1` = 1, `2` = 2))
surv.landings <- structure(
  bind_rows(read_csv('data_provided/surv.csv'),
            read_csv('data_provided/surv.csv') %>% 
              mutate(area=2)),
            area_group = list(`1` = 1, `2` = 2))

lln.landings  %>% 
  DT::datatable(.)  #this last line not necessary - helps with printing to web page
surv.landings  #%>% 
  DT::datatable(.)  #this last line not necessary - helps with printing to web page

```

Then when these catch data are included in the fleet files, the areas of operation 
with the fleets will be automatically updated. The current Rgadget version cannot 
update fleet components separately, so here we rewrite the entire fleet file the 
same way we did on day 2.

```{r lik2}

gadgetfleet('Modelfiles/fleet',gd,missingOkay = TRUE) %>% 
  gadget_update('totalfleet',
                livesonareas = c(1,2),
                name = 'surv',
                suitability = 
                  list(lingimm = list(type='function',suit_func='exponentiall50',
                                      alpha = '#ling.surv.alpha',
                                      beta = '#ling.surv.l50'),
                       lingmat = list(type='function',suit_func='exponentiall50',
                                      alpha = '#ling.surv.alpha',
                                      beta = '#ling.surv.l50')),
                data = surv.landings) %>%
  gadget_update('totalfleet',
                name = 'lln',
                livesonareas = c(1,2),
                suitability = 
                  list(lingimm = list(type='function',suit_func='exponentiall50',
                                      alpha = '#ling.lln.alpha',
                                      beta = '#ling.lln.l50'),
                       lingmat = list(type='function',suit_func='exponentiall50',
                                      alpha = '#ling.lln.alpha',
                                      beta = '#ling.lln.l50')),
                data = lln.landings) %>% 
  gadget_update('totalfleet',
                name = 'bmt',
                livesonareas = c(1,2),
                suitability = 
                  list(lingimm = list(type='function',suit_func='exponentiall50',
                                      alpha = '#ling.bmt.alpha',
                                      beta = '#ling.bmt.l50'),
                       lingmat = list(type='function',suit_func='exponentiall50',
                                      alpha = '#ling.bmt.alpha',
                                      beta = '#ling.bmt.l50')),
                data = bmt.landings) %>% 
  gadget_update('totalfleet',
                name = 'gil',
                livesonareas = c(1,2),
                suitability = 
                  list(lingimm = list(type='function',suit_func='exponentiall50',
                                      alpha = '#ling.gil.alpha',
                                      beta = '#ling.gil.l50'),
                       lingmat = list(type='function',suit_func='exponentiall50',
                                      alpha = '#ling.gil.alpha',
                                      beta = '#ling.gil.l50')),
                data = gil.landings) %>% 
  gadget_update('totalfleet',
                name = 'foreign',
                livesonareas = c(1,2),
                suitability = 
                  list(lingimm = list(type='function',suit_func='exponentiall50',
                                      alpha = '#ling.lln.alpha',
                                      beta = '#ling.lln.l50'),
                       lingmat = list(type='function',suit_func='exponentiall50',
                                      alpha = '#ling.lln.alpha',
                                      beta = '#ling.lln.l50')),
                data = for.landings) %>% 
  write.gadget.file(gd)
```

Now that there are fleets fishing on area 2 with catches removed, 
it is possible for biological samples to be generated from area 2. 

```{r, echo = FALSE}
#some necessary info
minage <- ling.imm[[1]]$minage
maxage <- ling.mat[[1]]$maxage
maxlength <- ling.mat[[1]]$maxlength 
minlength <- ling.imm[[1]]$minlength
dl <- ling.imm[[1]]$dl
slices <- 
  list('len20' = c(20,52), 'len52' = c(52,60), 'len60' = c(60,72), 'len72' = c(72,80), 'len80' = c(80,92), 'len92' = c(92,100), 'len100' = c(100,160) )

create_intervals <-  function (prefix, vect) {
  
    x<-
    structure(vect[1:(length(vect)-1)], names = paste0(prefix, vect[1:(length(vect)-1)])) %>% 
    as.list(.) %>% 
    purrr::map(~structure(., 
                          min = ., 
                          max = vect[-1][which(vect[1:(length(vect)-1)]==.)]))
    
    return(x)
}

#length distributions
ldist.surv <- 
  structure(bind_rows(read_csv('data_provided/ldist_sur.csv'),
                      read_csv('data_provided/ldist_sur.csv') %>% 
                        mutate(area = 2,
                               number = round(number*rlnorm(n(), sdlog = 0.5))) %>%
                        filter(number > 0)
                      ),
            area = as.list(area_file[[1]]$areas %>% set_names(.)),
            length = create_intervals('len',seq(minlength, maxlength, dl)),
            age = create_intervals('all',seq(minage, maxage, maxage-minage))
            )
ldist.lln <- 
  structure(bind_rows(read_csv('data_provided/ldist_lln.csv'),
                      read_csv('data_provided/ldist_lln.csv') %>% 
                        mutate(area = 2,
                               number = round(number*rlnorm(n(), sdlog = 0.5))) %>%
                        filter(number > 0)
                      ),
            area = as.list(area_file[[1]]$areas %>% set_names(.)),
            length = create_intervals('len',seq(minlength, maxlength, dl)),
            age = create_intervals('all',seq(minage, maxage, maxage-minage))
            )
ldist.bmt <- 
  structure(bind_rows(read_csv('data_provided/ldist_bmt.csv'),
                      read_csv('data_provided/ldist_bmt.csv') %>% 
                        mutate(area = 2,
                               number = round(number*rlnorm(n(), sdlog = 0.5))) %>%
                        filter(number > 0)
                      ),
            area = as.list(area_file[[1]]$areas %>% set_names(.)),
            length = create_intervals('len',seq(minlength, maxlength, dl)),
            age = create_intervals('all',seq(minage, maxage, maxage-minage))
            )
ldist.gil <- 
  structure(bind_rows(read_csv('data_provided/ldist_gil.csv'),
                      read_csv('data_provided/ldist_gil.csv') %>% 
                        mutate(area = 2,
                               number = round(number*rlnorm(n(), sdlog = 0.5))) %>%
                        filter(number > 0)
                      ),
            area = as.list(area_file[[1]]$areas %>% set_names(.)),
            length = create_intervals('len',seq(minlength, maxlength, dl)),
            age = create_intervals('all',seq(minage, maxage, maxage-minage))
            )

#age-length distributions
aldist.surv <- 
  structure(bind_rows(read_csv('data_provided/aldist_sur.csv'),
                      read_csv('data_provided/aldist_sur.csv') %>% 
                        mutate(area = 2,
                               number = round(number*rlnorm(n(), sdlog = 0.5))) %>%
                        filter(number > 0)
                      ),
            area = as.list(area_file[[1]]$areas %>% set_names(.)),
            length = create_intervals('len',seq(minlength, maxlength, dl)),
            age = create_intervals('age',c(minage:11, maxage))
            )
aldist.lln <- 
  structure(bind_rows(read_csv('data_provided/aldist_lln.csv'),
                      read_csv('data_provided/aldist_lln.csv') %>% 
                        mutate(area = 2,
                               number = round(number*rlnorm(n(), sdlog = 0.5))) %>%
                        filter(number > 0)
                      ),
            area = as.list(area_file[[1]]$areas %>% set_names(.)),
            length = create_intervals('len',seq(minlength, maxlength, dl)),
            age = create_intervals('age',c(minage:11, maxage))
            )
aldist.bmt <- 
  structure(bind_rows(read_csv('data_provided/aldist_bmt.csv'),
                      read_csv('data_provided/aldist_bmt.csv') %>% 
                        mutate(area = 2,
                               number = round(number*rlnorm(n(), sdlog = 0.5))) %>%
                        filter(number > 0)
                      ),
            area = as.list(area_file[[1]]$areas %>% set_names(.)),
            length = create_intervals('len',seq(minlength, maxlength, dl)),
            age = create_intervals('age',c(minage:11, maxage))
            )
aldist.gil <- 
  structure(bind_rows(read_csv('data_provided/aldist_gil.csv'),
                      read_csv('data_provided/aldist_gil.csv') %>% 
                        mutate(area = 2,
                               number = round(number*rlnorm(n(), sdlog = 0.5))) %>%
                        filter(number > 0)
                      ),
            area =as.list(area_file[[1]]$areas %>% set_names(.)),
            length = create_intervals('len',seq(minlength, maxlength, dl)),
            age = create_intervals('age',c(minage:11, maxage))
            )

#stock distribution data available:
matp.surv <- 
  structure(bind_rows(read_csv('data_provided/matp_sur.csv'),
                      read_csv('data_provided/matp_sur.csv') %>% 
                        mutate(area = 2,
                               number = round(number*rlnorm(n(), sdlog = 0.5))) %>%
                        filter(number > 0)
                      ),
            area = as.list(area_file[[1]]$areas %>% set_names(.)),
            length = create_intervals('len',seq(minlength, maxlength, dl*2)),
            age = list(mat_ages = ling.mat[[1]]$minage:ling.mat[[1]]$maxage)
            )

# set up aggregation info for length-based survey indices
SI1 <- 
  structure(bind_rows(read_csv('data_provided/SI1.csv'),
                      read_csv('data_provided/SI1.csv') %>% 
                        mutate(area = 2,
                               number = round(number*rlnorm(n(), sdlog = 0.5))/100) %>%
                        filter(number > 0)
                      ),
            area = as.list(area_file[[1]]$areas %>% set_names(.)),
            length = create_intervals('len', slices[[1]]))
SI2 <- 
  structure(bind_rows(read_csv('data_provided/SI2.csv'),
                      read_csv('data_provided/SI2.csv') %>% 
                        mutate(area = 2,
                               number = round(number*rlnorm(n(), sdlog = 0.5))/100) %>%
                        filter(number > 0)
                      ),
            area = as.list(area_file[[1]]$areas %>% set_names(.)),
            length = create_intervals('len', slices[[2]]))
SI3 <- 
  structure(bind_rows(read_csv('data_provided/SI3.csv'),
                      read_csv('data_provided/SI3.csv') %>% 
                        mutate(area = 2,
                               number = round(number*rlnorm(n(), sdlog = 0.5))/100) %>%
                        filter(number > 0)
                      ),
            area = as.list(area_file[[1]]$areas %>% set_names(.)),
            length = create_intervals('len', slices[[3]]))
SI4 <- 
  structure(bind_rows(read_csv('data_provided/SI4.csv'),
                      read_csv('data_provided/SI4.csv') %>% 
                        mutate(area = 2,
                               number = round(number*rlnorm(n(), sdlog = 0.5))/100) %>%
                        filter(number > 0)
                      ),
            area = as.list(area_file[[1]]$areas %>% set_names(.)),
            length = create_intervals('len', slices[[4]]))
SI5 <- 
  structure(bind_rows(read_csv('data_provided/SI5.csv'),
                      read_csv('data_provided/SI5.csv') %>% 
                        mutate(area = 2,
                               number = round(number*rlnorm(n(), sdlog = 0.5))/100) %>%
                        filter(number > 0)
                      ),
             area = as.list(area_file[[1]]$areas %>% set_names(.)),
            length = create_intervals('len', slices[[5]]))
SI6 <- 
  structure(bind_rows(read_csv('data_provided/SI6.csv'),
                      read_csv('data_provided/SI6.csv') %>% 
                        mutate(area = 2,
                               number = round(number*rlnorm(n(), sdlog = 0.5))/100) %>%
                        filter(number > 0)
                      ),
            area = as.list(area_file[[1]]$areas %>% set_names(.)),
            length = create_intervals('len', slices[[6]]))
SI7 <- 
  structure(bind_rows(read_csv('data_provided/SI7.csv'),
                      read_csv('data_provided/SI7.csv') %>% 
                        mutate(area = 2,
                               number = round(number*rlnorm(n(), sdlog = 0.5))/100) %>%
                        filter(number > 0)
                      ),
            area = as.list(area_file[[1]]$areas %>% set_names(.)),
            length = create_intervals('len', slices[[7]]))

```

Because all fish in the model, whether they belong to the area 1 or 2, belong to the
same stocks (immature or mature), they all need to conform to the same stock-related
processes, such as growth, maturation, and selectivity of a fleet as it fishes that
stock. Different stocks would be necessary to change these by area. However, because
initial conditions and recruitment are area-specific, they can be set with different values or 
switches. Previously we had written the same parameters for both areas, resulting in mirror
images; here we change these parameters to be area-specific. Also remember that
catchability is also estimated separately by area in relation to survey indices. 

```{r}
ling.imm <-
  ling.imm %>%
  gadget_update('initialconditions',
                normalparam = 
                  expand.grid(age = ling.imm[[1]]$minage:ling.imm[[1]]$maxage, area = area_file[[1]]$areas) %>% 
                  as_tibble() %>% 
                  mutate( 
                    age.factor = parse(text=sprintf('exp(-1*(lingimm.M+ling.init.F)*%1$s)*lingimm.init.%1$s',age)) %>% 
                      map(to.gadget.formulae) %>% 
                      unlist(),   
                    area.factor = sprintf('#lingimm.init.scalar.%1$s',area),
                    mean = von_b_formula(age,linf='ling.Linf',k='ling.k',recl='ling.recl'),
                    stddev = rep(init.sigma.i,2),
                    alpha = '#ling.walpha',
                    beta = '#ling.wbeta'
                    )
                ) %>%
  gadget_update('doesrenew',
                normalparam = 
                  expand.grid(year = schedule$year,area = area_file[[1]]$areas) %>%
                  distinct() %>% 
                  as_tibble() %>% 
                  mutate(step = 1,
                         age = ling.imm[[1]]$minage,
                         number = parse(text=sprintf('ling.rec.scalar*ling.rec.%1$s.%2$s',year,area)) %>% 
                           map(to.gadget.formulae) %>% 
                           unlist(),
                         mean = von_b_formula(age,linf='ling.Linf',k='ling.k',recl='ling.recl'),
                         stddev = '#ling.rec.sd',
                         alpha = '#ling.walpha',
                         beta = '#ling.wbeta'
                         )
                ) 

ling.mat <-
  ling.mat %>% 
  gadget_update('initialconditions',
                normalparam = expand.grid(age = ling.mat[[1]]$minage:ling.mat[[1]]$maxage, area = area_file[[1]]$areas) %>% 
                  as_tibble() %>% 
                  mutate(
                    age.factor = parse(text=sprintf('exp(-1*(lingmat.M+ling.init.F)*%1$s)*lingmat.init.%1$s',age)) %>% 
                      map(to.gadget.formulae) %>% 
                      unlist(),
                    area.factor = sprintf('#lingmat.init.scalar.%1$s',area),
                    mean = von_b_formula(age,linf='ling.Linf',k='ling.k',recl='ling.recl'),
                    stddev = rep(init.sigma.m,2),
                    alpha = '#lingmat.walpha',
                    beta = '#lingmat.wbeta'
                    )
                )  
## write to file
ling.imm %>% 
  write.gadget.file(gd)

ling.mat %>% 
  write.gadget.file(gd)

```


Finally, these data can be used to rewrite the likelihood function.


```{r}

lik <-
  gadgetlikelihood('likelihood',gd,missingOkay = TRUE) %>% 
  gadget_update("catchdistribution",
                name = "ldist.surv",
                weight = 1,
                data = ldist.surv,
                fleetnames = c("surv"),
                stocknames =c("lingimm","lingmat")) %>% 
  gadget_update("catchdistribution",
                name = "aldist.surv",
                weight = 1,
                data = aldist.surv, 
                fleetnames = c("surv"),
                stocknames =c("lingimm","lingmat")) %>% 
  gadget_update("catchdistribution",
                name = "ldist.lln",
                weight = 1,
                data = ldist.lln, 
                fleetnames = c("lln"),
                stocknames = c("lingimm","lingmat")) %>% 
  gadget_update("catchdistribution",
                name = "aldist.lln",
                weight = 1,
                  data = aldist.lln,
                fleetnames = c("lln"),
                stocknames = c("lingimm","lingmat")) %>% 
  gadget_update("catchdistribution",
                name = "ldist.gil",
                weight = 1,
                data = ldist.gil,
                fleetnames = c("gil"),
                stocknames = c("lingimm","lingmat")) %>% 
  gadget_update("catchdistribution",
                name = "aldist.gil",
                weight = 1,
                data = aldist.gil,
                fleetnames = c("gil"),
                stocknames = c("lingimm","lingmat")) %>% 
  gadget_update("catchdistribution",
                name = "ldist.bmt",
                weight = 1,
                data = ldist.bmt,
                fleetnames = c("bmt"),
                stocknames = c("lingimm","lingmat")) %>% 
  gadget_update("catchdistribution",
                name = "aldist.bmt",
                weight = 1,
                data = aldist.bmt,
                fleetnames = c("bmt"),
                stocknames = c("lingimm","lingmat")) %>% 
  gadget_update("stockdistribution",
                name = "matp.surv",
                weight = 1,
                  data = matp.surv,
                fleetnames = c("surv"),
                stocknames =c("lingimm","lingmat")) %>%  
  gadget_update("surveyindices",
                name = paste0("si.", names(slices)[1]),
                weight = 1,
                data = SI1,
                fittype = 'loglinearfit',
                stocknames = c("lingimm","lingmat")) %>% 
  gadget_update("surveyindices",
                name = paste0("si.", names(slices)[2]),
                weight = 1,
                data = SI2,
                fittype = 'loglinearfit',
                stocknames = c("lingimm","lingmat")) %>% 
  gadget_update("surveyindices",
                name = paste0("si.", names(slices)[3]),
                weight = 1,
                data = SI3,
                fittype = 'fixedslopeloglinearfit',
                slope=1,
                stocknames = c("lingimm","lingmat")) %>% 
  gadget_update("surveyindices",
                name = paste0("si.", names(slices)[4]),
                weight = 1,
                data = SI4,
                fittype = 'fixedslopeloglinearfit',
                slope=1,
                stocknames = c("lingimm","lingmat")) %>% 
  gadget_update("surveyindices",
                name = paste0("si.", names(slices)[5]),
                weight = 1,
                data = SI5,
                fittype = 'fixedslopeloglinearfit',
                slope=1,
                stocknames = c("lingimm","lingmat")) %>% 
  gadget_update("surveyindices",
                name = paste0("si.", names(slices)[6]),
                weight = 1,
                data = SI6,
                fittype = 'fixedslopeloglinearfit',
                slope=1,
                stocknames = c("lingimm","lingmat")) %>% 
  gadget_update("surveyindices",
                name = paste0("si.", names(slices)[7]),
                weight = 1,
                data = SI7,
                fittype = 'fixedslopeloglinearfit',
                slope=1,
                stocknames = c("lingimm","lingmat"))%>% 
  gadget_update("penalty",
                name = "bounds",
                weight = "0.5",
                data = data.frame(
                  switch = c("default"),
                  power = c(2),
                  upperW=10000,
                  lowerW=10000,
                  stringsAsFactors = FALSE)) %>%
  gadget_update("understocking",
                name = "understocking",
                weight = "100")

lik %>% 
  write.gadget.file(gd)
 
```

#### Exercise
* Now that the area model is set up with new parameters and area structure, 1) use `callGadget` 
to create a new parameters file, 2) use `init_guess` to set bounds and initial starting values 
of all parameters in the new parameters file, and 3) use `callGadget` again to take create 
another parameter file that contains reasonable starting values for a longer Gadget optimization 
run, and use 4) `gadget.reiterative` to create a final model run with decent
weights among data sources. If you are not sure where to start, look back at
the procedure in day 3.

* Now run gadget.fit on the results and create model fit figures that are area-specific.


```{r, echo = FALSE}
gadget_evaluate(gd,log = 'two_area_log')
lw_pars <- c(0.00000228, 3.20) #same values as in day 2

## update the input parameters with sane initial guesses
read.gadget.parameters(sprintf('%s/params.out',gd)) %>% 
  init_guess('rec.[0-9]|init.[0-9]',1,0.001,1000,1) %>%
  init_guess('recl',12,4,20,1) %>% 
  init_guess('rec.sd',5, 4, 20,1) %>% 
  init_guess('Linf',160, 100, 200,1) %>% 
  init_guess('k$',90, 40, 100,1) %>% 
  init_guess('bbin',6, 1e-08, 100, 1) %>% 
  init_guess('alpha', 0.5,  0.01, 3, 1) %>% 
  init_guess('l50',50,10,100,1) %>% 
  init_guess('walpha',lw_pars[1], 1e-10, 1,0) %>% 
  init_guess('wbeta',lw_pars[2], 2, 4,0) %>% 
  init_guess('M$',0.15,0.001,1,0) %>% 
  init_guess('rec.scalar',400,1,500,1) %>% 
  init_guess('init.scalar',200,1,300,1) %>% 
  init_guess('mat2',60,0.75*60,1.25*60,1) %>% 
  init_guess('mat1',70,  10, 200, 1) %>% 
  init_guess('init.F',0.4,0.1,1,1) %>% 
  init_guess('mig',0.1,0.1,1,0) %>% 
  write.gadget.parameters(.,file=sprintf('%s/params.in',gd))

read.gadget.parameters(sprintf('%s/params.in',gd))  #%>% 
 # DT::datatable(.)  #this last line not necessary - helps with printing to web page
gadget_optimize(gd,params.in ='params.in',params.out = 'params.init')

old <- getwd()
setwd(gd)

unlink("WGTS", recursive=TRUE)
timestamp() 
run <- gadget.iterative()
timestamp()


fit <- gadget.fit()
plot(fit)
tmp <- plot(fit,data = 'catchdist.fleets')
tmp$ldist.lln  %>% 
  DT::datatable(.)  #this last line not necessary - helps with printing to web page

```



