---
title: "Getting started"
output: 
  html_document:
    fig_height: 4
    fig_width: 8
    highlight: haddock
    theme: united
    toc: yes
    toc_float: yes
bibliography: library.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Install devtools, Rgadget and mfdb. Rgadget is under constant construction so please remember to update it regularly (e.g., monthly). 

```{r setupRgadget, eval=FALSE}
install.packages('devtools')
devtools::install_github('hafro/rgadget')
devtools::install_github('mareframe/mfdb')
```

Note that you can update all packages, both from github and CRAN using the `update_packages` command from `devtools`

## What is Rgadget?

Because it was originally designed as an ecosystem
simulator, Gadget already has a number of optional features structurally integrated 
with basic population dynamics, making it extremely flexible for representing a wide
variety of real-life biological scenarios (e.g., substock structure, length-based
processes, predator-prey dynamics, temperature-dependent growth). However, Gadget 
was created when Unix workstations were the most powerful and popular operating
systems available, and therefore lacks by itself a
user-friendly GUI. It is a collection of C++ scripts that comprise its own program,
similar to models created using ADMB. Running Gadget 
requires providing a rather large set of individual text files that specify model
settings and data. As a result, creating
more complex models involving a large set of similar input text files can become
tedious and error-prone, and can impair reproducability.

```{r engine='tikz',echo=FALSE}

\usetikzlibrary{arrows,decorations.pathmorphing,decorations.footprints,fadings,calc,trees,mindmap,shadows,decorations.text,patterns,positioning,shapes,matrix,fit,backgrounds}
 \begin{tikzpicture}[xscale = .9,yscale = .9]
    \node (main)[shape=ellipse,style=dashed,draw] at (-7,-2){main};
    \node (time) at (-2,2){time};
    \node (area) at (-2,1){area};
    \node (print) at (-2,0){printfile};
    \node (out)[shape=rectangle,draw] at (2.5,2) {output};
    \node (stock)[shape=rectangle,draw] at (-2,-1){stockfiles};
    \node (refw) at (4,0){reference weights};
    \node (initpop) at (4,-0.5){initial population};
    \node (rec) at (4,-1){recruitment};
    \node (mat) at (4,-1.5){maturation};
    \node (mig) at (4,-2){migration};
    \node (spw) at (4,-2.5){spawning};
    \node (str) at (4,-3){straying};
    \node (tagf)[shape=rectangle,draw] at (-2,-2.5){tagfiles};
    \node (tagd) at (2.5,-4){tag data};
    \node (oth)[shape=rectangle,draw] at (-2,-4){otherfoodfiles};
    \node (othd) at (2.5,-5){otherfood data};
    \node (fleet)[shape=rectangle,draw] at (-2,-5.5){fleetfiles};
    \node (fleetd) at (2.5,-6){fleet data};
    \node (lik)[shape=rectangle,draw] at (-2,-7){likelihoodfiles};
    \node (likd) at (2.5,-7){likelihood data};
   
    \node (param)[shape=ellipse,style=dashed,draw] at (-7,-4){parameters};
    \node (opt)[shape=ellipse,style=dashed,draw] at (-7,-7){optimisation};
    
    \draw[->] (main) to (time);
    \draw[->] (main) to (area);
    \draw[->] (main) to (print);
    \draw[->] (main) to (stock);
    \draw[->] (main) to (tagf);
    \draw[->] (main) to (oth);
    \draw[->] (main) to (fleet);
    \draw[->] (main) to (lik);
    
    \draw[->] (print) to (out);
    
    \draw[->] (stock) to (refw);
    \draw[->] (stock) to (initpop);
    \draw[->] (stock) to (rec);
    \draw[->] (stock) to (mat);
    \draw[->] (stock) to (mig);
    \draw[->] (stock) to (spw);
    \draw[->] (stock) to (str);
    
    \draw[->] (tagf) to (tagd);
    
    \draw[->] (oth) to (othd);
    \draw[->] (fleet) to (fleetd);
    \draw[->] (lik) to (likd);
    \end{tikzpicture}
```

Rgadget was intended to be a solution to this problem, by linking the power behind 
Gadget with the familiar R interface. Therefore, 90% of Rgadget's functionality is file 
and data handling: it creates files that are in a correct format so that Gadget understands 
them, creates organized directories for storing those files, takes 
care of main file dependencies and compatibility issues, and creates a
direct pipeline for inserting data into such files. It additionally runs Gadget,
implements an iterative reweighting algorithm to aid with data weighting issues when
there are multiple sources of data, reads file output, and merges input and output data
so that meaningful illustrative and diagnostic figures can be made (a variety of which
come predefined). Rgadget is your best friend if you want to quickly make reproducable,
trackable, interpretable, and sharable models using Gadget.

Gadget has a variety of convenience functions included that will become more visible
as examples are walked through, but the main types of functions to recognize include:

1. file creation functions (e.g, gadgetfile, gadgetstock, gadgettime, gadgetfleet,
gadgetlikelihood, gadgetdata, etc.), 
2. file I/O functions (e.g.,write.gadget.file, read.gadget.file, etc.), 
3. file modification functions (e.g., init_guess, gadget_update), 
4. functions that run Gadget (e.g., callGadget, gadget.iterative, gadget.forward), 
5. functions that compile input and output and create plots to analyze model fits 
(e.g., gadget.fit, bind.gadget.fit).

Using this set of functions, the general work flow for using Gadget to fit a statistical
model via Rgadget is demonstrated in our examples and follows the pattern:

1. Create new Gadget input file objects within R and write them to disk to comprise an
initial Gadget model structure.
2. Run an initial Gadget simulation to test files and conveniently create input
paramater files. 
3. Read certain Gadget model files back into R, so that they can be modified and then
overwritten. 
4. Run an initial Gadget fitting procedure to conveniently create more input files
(i.e., decent starting values).
5. Set optimisation function parameters and run a statistical Gadget model run using
reiterative weighting to determine relative contribution of each data source to the
objective function.
6. Visualize the model fit.

As most programming errors and debugging occur in steps 1 - 4, most of this course will
focus on these steps. Statistical model fitting and visualization will be demonstrated,
but details regarding the statistical theory behind fitting procedures, analysis of
model fit, and diagnostics are beyond the scope of this course.

## Using Rgadget
To use the Rgadget functions you will need to load it into your R workspace in the usual manner:
```{r message=FALSE,warning=FALSE}
library(Rgadget)
```
For convenience Rgadget will also load the tidyverse package as it is used extensively 
when interacting with Gadget and viewing the output. As noted above Rgadget has a number
of functions and we will introduce the main functions in Rgadget as we go through 
the tutorial. 

Here we will focus on setting up a simple dummy model using the `gadgetfile` function and
functions that are its derivatives. The `gadgetfile` function is low level function that
creates an R object that can be exported to files that coheres to the Gadget file format,
described in the [Userguide](https://hafro.github.io/gadget/docs/userguide). Essentially a
Gadget file is structured by a list of named components and within those components there
are named settings. As an example of this you can create your own custom Gadget file by
specifying the components as named lists embedded within a named list:
```{r}
dummy <- 
  gadgetfile('test',file_type = 'generic', 
             components = list(fruit=list(apple=1,orange=5),
                               vegetable=list(cucumber=1)))

dummy
```
You will note that when calling the `gadgetfile` object the printed output will be in
the same format as the resulting Gadget file. This for our mainly convenience as you 
will be able to directly see if the format is incorrect, by comparing it to the format
described in the [Userguide](https://hafro.github.io/gadget/docs/userguide), before
writing it to file. If you need however to access individual components of you can of
course do so by calling the name of the component directly:

```{r}
dummy$fruit
```

and if you want to investigate the structure further you can call `str` on the object:
```{r}
str(dummy)
```

The file type argument in `gadgetfile` allows the function to customize the output
depending to the various quirks of the file. You can see the different file types
currently allowed for by `gadgetfile` by calling `gadget_filetypes`:
```{r echo=FALSE}
gadget_filetypes %>% 
  knitr::kable()
```

The output can then subsequently be written to a file using `write.gadget.file`:
```{r eval=FALSE}
write.gadget.file(obj=dummy,path='somewhere_over_the_rainbow')
```


## Creating model input files 

Having described how `gadgetfile` works we can now start to define our first model, 
which we will refer to as the "Simple model". This is done by first deciding where
to put the model and the associated files. Rgadget helps you by creating a convenience
function `gadget.variant.dir` whichs set up a working directory for the Gadget model:

```{r}
gd <- gadget.variant.dir('simple_model')
gd
```

You will notice that `gd` is just a simple string with the location of the model, which
in this case is a folder named `simple_model` within the current working directory, and
an attribute to that string called `mainfile`. The main file is, as illustrated below, 
the root of the Gadget input file structure, specifiying all the paths to the definition
of the ecosystem that is being simulated, such as the stocks and time period. By invoking
the `gadget.variant.dir` command a new folder for the model is created and subsequent
edits to the `gd` object will update the main file as necessary. 

```{r engine='tikz',echo=FALSE}
\usetikzlibrary{arrows,decorations.pathmorphing,decorations.footprints,fadings,calc,trees,mindmap,shadows,decorations.text,patterns,positioning,shapes,matrix,fit,backgrounds}
 \begin{tikzpicture}[xscale = 1.5,yscale = 1]
    \node (main)[shape=ellipse,style=dashed,draw] at (0,0){main};
    \node (time) at (-2,2){time};
    \node (area) at (-2,1){area};
    \node (print)[color=gray] at (-2,0){printfile};
    \node (stock)[shape=rectangle,draw] at (-2,-1){stockfiles};
    \node (tagf)[shape=rectangle,draw,color=gray] at (2,2){tagfiles};
    \node (oth)[shape=rectangle,draw,color=gray] at (2,1){otherfoodfiles};
    \node (fleet)[shape=rectangle,draw] at (2,0){fleetfiles};
    \node (lik)[shape=rectangle,draw,color=gray] at (2,-1){likelihoodfiles};


    \draw[->] (main) to (time);
    \draw[->] (main) to (area);
    \draw[->] (main) to (print);
    \draw[->] (main) to (stock);
    \draw[->] (main) to (tagf);
    \draw[->] (main) to (oth);
    \draw[->] (main) to (fleet);
    \draw[->] (main) to (lik);
    \end{tikzpicture}
```

We will now go through each of these different file types and show you how to generate
them. Note that the grayed files in the illustration above will be covered at a later
stage. 

### Time file
Gadget is structured as a forward simulator, so in order for it to run we will first need
to define the time period ('year'), and number of subannual time steps ('step'), for the
projection. We do so by creating a data frame setting up the schedule for the simulation:

```{r}
schedule <- 
  expand.grid(year = 1:10, step = 1:4) %>% 
  arrange(year)
```

and create a new `time` file using `gadgetfile` by explicitly setting the file type as
'time':

```{r}
## think about writing a special function for this using the schedule data.frame

gadgetfile('Modelfiles/time',
           file_type = 'time',
           components = list(list(firstyear = min(schedule$year),
                                  firststep=1,
                                  lastyear=max(schedule$year),
                                  laststep=4,
                                  notimesteps=c(4,3,3,3,3)))) %>% ## number of time steps and step length in months
  write.gadget.file(gd)
```
Note that arbitrary years and timesteps can be defined for the model, as appropriate for
 the data and the species simulated. One can for instance start the simulation in 1981 to 2019, with four timesteps with a timestep length of 1,2,4 and 5 months. Typically, though, Gadget models tend to be set up in a set number of year with quarterly timesteps. 

### Area file

Since Gadget allows for simulations in multiple areas these need to be explicitly defined.
Various processes, such as fleet operations and migrations, in the model can then be
explicitly defined by area. As for the time file the areas are defined in the area file
which you can create using `gadgetfile`:

```{r}
gadgetfile('Modelfiles/area',
           file_type = 'area',
           components = list(list(areas = 1,
                                  size = 1,
                                  temperature = schedule %>% 
                                    mutate(area = 1, temperature = 5)))) %>% 
  write.gadget.file(gd)
```
which will generate a file starting with:
```{r echo=FALSE}
gadgetfile('Modelfiles/area',
           file_type = 'area',
           components = list(list(areas = 1,
                                  size = 1,
                                  temperature = schedule %>% head(5) %>%  
                                    mutate(area = 1, temperature = 5))))
```

The areas are defined with attributes, size of the area and average temperature for each
time step. Note although these attributes are rarely used in practice they will need to be
defined for all areas and, in the case of temperatrue, time steps in order for the
simulation to run. 

### Stockfiles

Rgadget has a number of convenience functions when dealing with more complicated files. These functions are all derivatives of `gadgetfile` and thus can be written to file using `write.gadget.file`. Here we will start off by creating a skeleton stock file using the `gadgetstock` function:

```{r}
stock <- gadgetstock('simple_stock',gd,missingOkay = TRUE)
stock
```
This will define a stock named "simple_stock" within the "simple_model" ecosystem. This
stock file is however not a complete file so we will need to edit the `stock` object by
adding the necessary attributes. This is done using `gadget_update` which, as the name
suggests, allows to you to update individual components of the file. The first argument to
`gadget_update` after the stock file is the name of the component you want update followed
by the variables within the component you want to update. So let's start by setting
mininum and maximum age and other similar attributes in the stock component:

```{r}
  stock %>% 
  gadget_update('stock',
                livesonareas = 1,
                maxage = 1,
                minage = 1,
                minlength = 0,
                maxlength = 2,
                dl = 1)
```

As `gadget_update` was designed to allow the user to edit the Gadget file using a piping
mechanism, a la tidyverse, we can of course do multiple edits using `gadget_update` in one
go:

```{r}
stock <- 
  stock %>% 
  gadget_update('stock',
                livesonareas = 1,
                maxage = 1,
                minage = 1,
                minlength = 0,
                maxlength = 2,
                dl = 1) %>% 
  gadget_update('doesgrow',0) %>% 
  gadget_update('naturalmortality',0) %>% 
  gadget_update('refweight',data=tibble(length=0:2,mean=0:2)) %>% 
  gadget_update('initialconditions',
                normalparam = tibble(age = 1,
                                     area = 1,
                                     age.factor = 1,   
                                     area.factor =1,
                                     mean = 1,
                                     stddev = .1,
                                     alpha = 1,
                                     beta = 1))
stock %>% 
  write.gadget.file(gd)
stock
```

There is a lot to digest from the code above so it is worth stopping and taking a look at each element: 

* The first call to `gadget_update` edited the attributes of the "stock" part of the file as discussed above. 
* The second call to "doesgrow" turned off all growth processes in the model for this stock by setting `doesgrow` to `0`. (Note that we will discuss growth models in later chapters). 
* Natural mortality was set 0, meaning that fish in the model will not die of natural causes.  
* The weight of the fish was defined. 
* The initial number of fish and their size range at age was defined via "initalconditions". The initial number of fish is determined in multiples of 10 000 
fish, so to derive the number of fish at a certain age and in a specific area you need 
to multiply the `area.factor` with the `age.factor`. With in the same cell you can r
ead the mean length and standard deviation from the `mean` and `stddev` columns, and 
the paramters of the weight--length relationship ($w = \alpha l^\beta$) can be read 
from the `alpha` and `beta` columns. 


The stock file essentially defines a stock with 10 000 fish that do not grow, consume 
or die for the duration of the simulation. By writing this file to the disk R has 
behind the scenes created a folder named `simple_model` with all the necessary files:
```{r}
fs::dir_ls(gd)
```


## Starting a simulation

To test if this is working we can run a single simulation first to check for any errors
or warning in our setup. 

```{r}
## hide the setenv call in callGadget or create a nicer frontend to callGadget
Sys.setenv(GADGET_WORKING_DIR=normalizePath(gd)) 
callGadget(s=1,main='main',ignore.stderr = FALSE,log = 'simple_log')
```

This gives you a warning that no understocking likelihood is available. The 
understocking likelihood is a penalty function used to prevent negative consumption, 
i.e. more fish being eaten and/or fished than is available. We can safely ignore this
warning for the time being. If you get additional errors, ending with 


```
...running command 'gadget  -s      -main main      -log simple_log   ' had status 1
```

your Gadget run did
not work. At this stage in the tutorial, it is likely becuase you have previously run 
later code from the next section, which has provided additional old files to your 
Gadget model directory (see above) that do not match the specifications above. 
This can be easily fixed by deleting the 'simple_log' directory and then rerunning 
the code on this page from the point of defining `gd`.

After running Gadget we see that two new files have been 
added to the directory, the parameter file (`params.out`) and the log file 
(`simple_log`). We will discuss the parameter file in more detail in the following
sections but for now lets look at the log file: 

**If and when you have errors, such as the '...status 1' error above, the log file is the best place to start trying to figure out why the error occurs.** 

```{r}
read_lines(paste(gd,'simple_log',sep='/'))[-(58:175)] 
```


We can now gather the output from the model using `gadget.fit`:
```{r}
fit <- gadget.fit(wgts = NULL,gd = gd)
plot(fit,data='res.by.year',type = 'num.total') 
```
and the fit object contains a number of useful datasets that can be used for further
analysis:
```{r}
names(fit)
```


