---
title: "Getting started"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting started

Install devtools and Rgadget. Rgadget is under constant construction so please remember to update it regularly (e.g., monthly).

```{r}
install.packages('devtools')
devtools::install_github('hafro/rgadget')
library(Rgadget)
#library(mfdb)
```

# What is Rgadget?

Gadget was created long before R or even Windows (?), and therefore lacks by itself a
user-friendly GUI. It is a collection of C++ scripts that comprise its own program,
similar to how models created using ADMB function. However, running a Gadget 
requires providing a rather large set of individual text files that specify model
settings and data. In addition, because it was originally designed as an ecosystem
simulator, Gadget already has a number of optional features structurally integrated 
with basic population dynamics, making it extremely flexible for representing a wide
variety of real-life biological scenarios (e.g., substock structure, length-based
processes, predator-prey dynamics, temperature-dependent growth). As a result, creating
more complex models involving a large set of similarly input text files can become
tedious and error-prone, and can impair reproducability.

Rgadget was intended to be a solution to this problem, by linking the power behind 
Gadget with the familiar R interface. Therefore, 90% of Rgadget's functionality is file 
and data handling: it creates files that are in the correct format for Gadget to
understand them, creates directories for storing those files in organizes sense, takes 
care of most file dependencies and compatibility issues between files, and creates a
direct way to input data into such files. It additionally runs Gadget, implements an
iterative reweighting algorithm to aid with data weighting issues when there are 
multiple sources of data, reads file output, and merges input and output data so that 
meaningful illustrative and diagnostic figures can be made (a variety of which come 
predefined). Rgadget is your friend.

# Using Rgadget

Set up a Gadget directory:

```{r}
gd <- gadget_directory('simple_model')
gd
```

You will notice that `gd` is just a simple list with the location of the model and what is called the `mainfile`. The main file is the root of the Gadget input file structure, specifiying all the paths to the definition of the ecosystem that is being simulated, such as the stocks and time period. 

```{r engine='tikz', echo=FALSE}
\usetikzlibrary{arrows,decorations.pathmorphing,decorations.footprints,
fadings,calc,trees,mindmap,shadows,decorations.text,patterns,positioning,shapes,matrix,fit,backgrounds}
 \begin{tikzpicture}[xscale = .9,yscale = .9]
    \node (main)[shape=ellipse,style=dashed,draw] at (-7,-2){main};
    \node (time) at (-2,2){time};
    \node (area) at (-2,1){area};
    \node (print) at (-2,0){printfile};
    \node (out)[shape=rectangle,draw] at (2.5,2) {output};
    \node (stock)[GenericNodeStyle] at (-2,-1){stockfiles};
    \node (refw) at (4,0){reference weights};
    \node (initpop) at (4,-0.5){initial population};
    \node (rec) at (4,-1){recruitment};
    \node (mat) at (4,-1.5){maturation};
    \node (mig) at (4,-2){migration};
    \node (spw) at (4,-2.5){spawning};
    \node (str) at (4,-3){straying};
    \node (tagf)[GenericNodeStyle] at (-2,-2.5){tagfiles};
    \node (tagd) at (2.5,-4){tag data};
    \node (oth)[GenericNodeStyle] at (-2,-4){otherfoodfiles};
    \node (othd) at (2.5,-5){otherfood data};
    \node (fleet)[GenericNodeStyle] at (-2,-5.5){fleetfiles};
    \node (fleetd) at (2.5,-6){fleet data};
    \node (lik)[GenericNodeStyle] at (-2,-7){likelihoodfiles};
    \node (likd) at (2.5,-7){likelihood data};
   
    \node (param)[shape=ellipse,style=dashed,draw] at (-7,-4){parameters};
    \node (opt)[shape=ellipse,style=dashed,draw] at (-7,-7){optimisation};
    
    \draw[->] (main) to (time);
    \draw[->] (main) to (area);
    \draw[->] (main) to (print);
    \draw[->] (main) to (stock);
    \draw[->] (main) to (tagf);
    \draw[->] (main) to (oth);
    \draw[->] (main) to (fleet);
    \draw[->] (main) to (lik);
    
    \draw[->] (print) to (out);
    
    \draw[->] (stock) to (refw);
    \draw[->] (stock) to (initpop);
    \draw[->] (stock) to (rec);
    \draw[->] (stock) to (mat);
    \draw[->] (stock) to (mig);
    \draw[->] (stock) to (spw);
    \draw[->] (stock) to (str);
    
    \draw[->] (tagf) to (tagd);
    
    \draw[->] (oth) to (othd);
    \draw[->] (fleet) to (fleetd);
    \draw[->] (lik) to (likd);
    \end{tikzpicture}
```

By invoking the `gadget_directory` command a new folder for the model is created and subsequent edits to the `gd` object will update the main file as necessary. 

First we will define the time period, and number of intervals, for the projection. We do so by creating a data frame setting up the schedule for the simulation:

```{r}
schedule <- 
  expand.grid(year = 1:10, step = 1:4) 
```

and create a new `time` file:

```{r}
gadgetfile('Modelfiles/time',
           file_type = 'time',
           components = list(list(firstyear = min(schedule$year),
                                  firststep=1,
                                  lastyear=max(schedule$year),
                                  laststep=4,
                                  notimesteps=c(4,3,3,3,3)))) %>% 
  write.gadget.file(gd$dir)
```


areas on which the model will operate. 