---
title: "Getting started"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting started

```{r}
library(Rgadget)
library(mfdb)
```


Set up a Gadget directory:

```{r}
gd <- gadget_directory('simple_model')
gd
```

You will notice that `gd` is just a simple list with the location of the model and what is called the `mainfile`. The main file the root of the Gadget input file structure, specifiying all the paths to the defintition of the ecosystem that is being simulated, such as the stocks and time period. 

```{r engine='tikz', echo=FALSE}
\usetikzlibrary{arrows,decorations.pathmorphing,decorations.footprints,
fadings,calc,trees,mindmap,shadows,decorations.text,patterns,positioning,shapes,matrix,fit,backgrounds}
 \begin{tikzpicture}[xscale = .9,yscale = .9]
    \node (main)[shape=ellipse,style=dashed,draw] at (-7,-2){main};
    \node (time) at (-2,2){time};
    \node (area) at (-2,1){area};
    \node (print) at (-2,0){printfile};
    \node (out)[shape=rectangle,draw] at (2.5,2) {output};
    \node (stock)[GenericNodeStyle] at (-2,-1){stockfiles};
    \node (refw) at (4,0){reference weights};
    \node (initpop) at (4,-0.5){initial population};
    \node (rec) at (4,-1){recruitment};
    \node (mat) at (4,-1.5){maturation};
    \node (mig) at (4,-2){migration};
    \node (spw) at (4,-2.5){spawning};
    \node (str) at (4,-3){straying};
    \node (tagf)[GenericNodeStyle] at (-2,-2.5){tagfiles};
    \node (tagd) at (2.5,-4){tag data};
    \node (oth)[GenericNodeStyle] at (-2,-4){otherfoodfiles};
    \node (othd) at (2.5,-5){otherfood data};
    \node (fleet)[GenericNodeStyle] at (-2,-5.5){fleetfiles};
    \node (fleetd) at (2.5,-6){fleet data};
    \node (lik)[GenericNodeStyle] at (-2,-7){likelihoodfiles};
    \node (likd) at (2.5,-7){likelihood data};
   
    \node (param)[shape=ellipse,style=dashed,draw] at (-7,-4){parameters};
    \node (opt)[shape=ellipse,style=dashed,draw] at (-7,-7){optimisation};
    
    \draw[->] (main) to (time);
    \draw[->] (main) to (area);
    \draw[->] (main) to (print);
    \draw[->] (main) to (stock);
    \draw[->] (main) to (tagf);
    \draw[->] (main) to (oth);
    \draw[->] (main) to (fleet);
    \draw[->] (main) to (lik);
    
    \draw[->] (print) to (out);
    
    \draw[->] (stock) to (refw);
    \draw[->] (stock) to (initpop);
    \draw[->] (stock) to (rec);
    \draw[->] (stock) to (mat);
    \draw[->] (stock) to (mig);
    \draw[->] (stock) to (spw);
    \draw[->] (stock) to (str);
    
    \draw[->] (tagf) to (tagd);
    
    \draw[->] (oth) to (othd);
    \draw[->] (fleet) to (fleetd);
    \draw[->] (lik) to (likd);
    \end{tikzpicture}
```

By invoking the `gadget_directory` command a new folder for the model is created and subsequent edits to the `gd` object will update the main file as necessary. 

First we will define the time period, and number of intervals, for the projection. We do so by creating a data frame setting up the schedule for the simulation:

```{r}
schedule <- 
  expand.grid(year = 1:10, step = 1:4) 
```

and create a new `time` file:

```{r}
gadgetfile('Modelfiles/time',
           file_type = 'time',
           components = list(list(firstyear = min(schedule$year),
                                  firststep=1,
                                  lastyear=max(schedule$year),
                                  laststep=4,
                                  notimesteps=c(4,3,3,3,3)))) %>% 
  write.gadget.file(gd$dir)
```


areas on which the model will operate. 