---
title: "Getting started"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting started

Install devtools and Rgadget. Rgadget is under constant construction so please remember to update it regularly (e.g., monthly).

```{r}
install.packages('devtools')
devtools::install_github('hafro/rgadget')
library(Rgadget)
#library(mfdb)
```

# What is Rgadget?

Gadget was created when Unix workstations were the most powerful and popular operating
systems available, and therefore lacks by itself a
user-friendly GUI. It is a collection of C++ scripts that comprise its own program,
similar to how models created using ADMB function. However, running a Gadget 
requires providing a rather large set of individual text files that specify model
settings and data. In addition, because it was originally designed as an ecosystem
simulator, Gadget already has a number of optional features structurally integrated 
with basic population dynamics, making it extremely flexible for representing a wide
variety of real-life biological scenarios (e.g., substock structure, length-based
processes, predator-prey dynamics, temperature-dependent growth). As a result, creating
more complex models involving a large set of similarly input text files can become
tedious and error-prone, and can impair reproducability.

Rgadget was intended to be a solution to this problem, by linking the power behind 
Gadget with the familiar R interface. Therefore, 90% of Rgadget's functionality is file 
and data handling: it creates files that are in a correct format that Gadget understands them, creates directories for storing those files in organizes sense, takes 
care of main file dependencies and compatibility issues among files, and creates a
direct pipeline for inserting data into such files. It additionally runs Gadget,
implements aniterative reweighting algorithm to aid with data weighting issues when
there are multiple sources of data, reads file output, and merges input and output data
so that meaningful illustrative and diagnostic figures can be made (a variety of which
comepredefined). Rgadget is your best friend if you want to quickly make reproducable,
trackable, and sharable models using Gadget.

Gadget has a variety of convenience functions included that will become more visible
as examples are walked through, but the main types of functions to recognize include:

1. file creation functions (e.g, gadgetstock, gadgettime, gadgetfleet,
gadgetlikelihood, gadgetfile, gadgetdata, etc.), 
2. file I/O functions (e.g.,write.gadget.file, read.gadget.file, read.gadget.data,
read.gadget.likelihood, etc.), 
3. file modification functions (e.g., init_guess, gadget_update), 
4. functions that run Gadget (e.g., callGadget, gadget.iterative, gadget.forward), 
5. functions that compile input and output and create plots to analyze model fits 
(e.g., gadget.fit, bind.gadget.fit).

Using this set of functions, the general work flow for using Gadget to fit a statistical
model via Rgadget is demonstrated in our examples and follows the pattern:

1. Create new Gadget input file objects within R and write them to disk to comprise an
initial Gadget model structure.
2. Run an initial Gadget simulation to test files and conveniently create input
paramater files. 
3. Read certain Gadget model files back into R, so that they can be modified and then
overwritten. 
4. Run an initial Gadget fitting procedure to conveniently create more input files
(i.e., decent starting values).
5. Set optimisation function parameters and run a statistical Gadget model run using
reiteraive weighting to determine relative contribution of each data source to the
objective function.
6. Visualize the model fit.

As most programming errors and debugging occur in steps 1 - 4, 90% of this course will
focus on these steps. Statistical model fitting and visualization will be demonstrated,
but details regarding the statistical theory behind fitting procedures, analysis of
model fit, and diagnostics are beyond the scope of this course.

# Using Rgadget

## Create new Gadget model input files 

Set up a Gadget directory:

```{r}
gd <- gadget_directory('simple_model')
gd
```

You will notice that `gd` is just a simple list with the location of the model and what is called the `mainfile`. The main file is the root of the Gadget input file structure, specifiying all the paths to the definition of the ecosystem that is being simulated, such as the stocks and time period. 

```{r engine='tikz', echo=FALSE}
\usetikzlibrary{arrows,decorations.pathmorphing,decorations.footprints,
fadings,calc,trees,mindmap,shadows,decorations.text,patterns,positioning,shapes,matrix,fit,backgrounds}
 \begin{tikzpicture}[xscale = .9,yscale = .9]
    \node (main)[shape=ellipse,style=dashed,draw] at (-7,-2){main};
    \node (time) at (-2,2){time};
    \node (area) at (-2,1){area};
    \node (print) at (-2,0){printfile};
    \node (out)[shape=rectangle,draw] at (2.5,2) {output};
    \node (stock)[GenericNodeStyle] at (-2,-1){stockfiles};
    \node (refw) at (4,0){reference weights};
    \node (initpop) at (4,-0.5){initial population};
    \node (rec) at (4,-1){recruitment};
    \node (mat) at (4,-1.5){maturation};
    \node (mig) at (4,-2){migration};
    \node (spw) at (4,-2.5){spawning};
    \node (str) at (4,-3){straying};
    \node (tagf)[GenericNodeStyle] at (-2,-2.5){tagfiles};
    \node (tagd) at (2.5,-4){tag data};
    \node (oth)[GenericNodeStyle] at (-2,-4){otherfoodfiles};
    \node (othd) at (2.5,-5){otherfood data};
    \node (fleet)[GenericNodeStyle] at (-2,-5.5){fleetfiles};
    \node (fleetd) at (2.5,-6){fleet data};
    \node (lik)[GenericNodeStyle] at (-2,-7){likelihoodfiles};
    \node (likd) at (2.5,-7){likelihood data};
   
    \node (param)[shape=ellipse,style=dashed,draw] at (-7,-4){parameters};
    \node (opt)[shape=ellipse,style=dashed,draw] at (-7,-7){optimisation};
    
    \draw[->] (main) to (time);
    \draw[->] (main) to (area);
    \draw[->] (main) to (print);
    \draw[->] (main) to (stock);
    \draw[->] (main) to (tagf);
    \draw[->] (main) to (oth);
    \draw[->] (main) to (fleet);
    \draw[->] (main) to (lik);
    
    \draw[->] (print) to (out);
    
    \draw[->] (stock) to (refw);
    \draw[->] (stock) to (initpop);
    \draw[->] (stock) to (rec);
    \draw[->] (stock) to (mat);
    \draw[->] (stock) to (mig);
    \draw[->] (stock) to (spw);
    \draw[->] (stock) to (str);
    
    \draw[->] (tagf) to (tagd);
    
    \draw[->] (oth) to (othd);
    \draw[->] (fleet) to (fleetd);
    \draw[->] (lik) to (likd);
    \end{tikzpicture}
```

By invoking the `gadget_directory` command a new folder for the model is created and subsequent edits to the `gd` object will update the main file as necessary. 

First we will define the time period, and number of intervals, for the projection. We do so by creating a data frame setting up the schedule for the simulation:

```{r}
schedule <- 
  expand.grid(year = 1:10, step = 1:4) 
```

and create a new `time` file:

```{r}
gadgetfile('Modelfiles/time',
           file_type = 'time',
           components = list(list(firstyear = min(schedule$year),
                                  firststep=1,
                                  lastyear=max(schedule$year),
                                  laststep=4,
                                  notimesteps=c(4,3,3,3,3)))) %>% 
  write.gadget.file(gd$dir)
```


areas on which the model will operate. ]

## Run initial Gadget simulation




This section has demonstrated a simple implementation of steps 1 - 2. A more complex 
implementation using all steps will be presented in the next sections of the course.

## Modify some Gadget model files

## Fit initial Gadget statistical model to generate starting values

## Set optimisation function parameters and fit Gadget statistical model

## Visualize Gadget statistical model fit

