---
title: "Introduction to Gadget"
output: html_document
bibliography: library.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# General context of Gadget

Ecosystem considerations appear to be an obvious, and necessary part,
of the management of marine resources. Several international
organizations have stressed the importance of studying ecosystems
interactions in terms of management [eg. @Sinclair2002255,
  @nammco1997,@iwc2002]. Optimum fisheries management based on ecosystem
consideration is however an elusive target. Even if the model
sufficiently describes the target ecosystem, @voss2014regional
illustrated that the definition of the this optimum is dependent on
the manager's view of the resource. Although the rationale for
ecosystem based approach to fisheries management is strong its
adoption has been slow, which @mollmann2013implementing suggests
is due to lack of strategy. However the effects of species
interactions on the management of a commercially exploited stock is a fairly
complicated and data intensive excercise, as it requires data not
directly observed by fisheries.  @dickey2013ecosystem note
therefore that the state of knowledge is not sufficient to embark on 
ecosystem engineering, although in theory this would have desirable effects
on the status of key resources, as such an undertaking may have
undesirable and unforseen consequences.
 
The choice of a modelling approach used to assimilate biological
information such as those obtained from scientific surveys also needs
some consideration. There are a number of approaches to ecosystems
models. According to @plaganyi2007models they broadly fall broadly into
four categories:
* Whole ecosystem approaches, where all trophic levels of the
  ecosystem are modelled [e.g. @christensen2004ecopath].
* Dynamic systems models, where where the whole ecosystem and
  possibly its effects to coastal communities is modelled
  [e.g. @link2010northeast].
* Extensions of single--species assesment models
  [e.g. @livingston1998incorporation].
* Minimum realistic models, where only a few but significant
  interactions are included. Examples of cases are models built using
  the Gadget framework [@begley2004overview].

Obviously these approaches differ substantiantly both in scope and
objectives. Minimimum realistic models and singles species extensions
are, as the names suggest, more focussed on answering key questions
and their parameters are estimated statistically. Modelling
approaches, like those using Atlantis [@fulton2005ecological] or
Ecopath with Ecosim [@christensen2004ecopath], often try to
include a larger part of the ecosystem where it becomes harder to
estimate the model parameters through comparisons with real
data. These grand models tend to be more useful when answering
strategic questions, such as discussed by @fulton2008atlantis,
rather than more tactical issues such as direct advice on next years
quota.


## Gadget -- a toolbox for fisheries stock assessments

Modelling approaches such as those allowed by Atlantis have, although
interesting, tended to have quite obvious drawbacks in terms of
comparisons to real data. Estimates of process parameters are far from
trivial in this setting, both in terms of available data and more
importantly computation time. A fullblown Atlantis model will
generally take a substantial time for a single simulation and as such
not obvious how one would obtain an ``optimal'' fit in a statistical
sense. That said, these types of model can provide valuable insight
into how an ecosystems would behave. In particular one could study
certain performance metrics, such as ecosystem indicators [as
described in @fulton2005ecological], and various stock assessment
models.

It has been observed that in various settings single species models of
a marine resource fail to explain model deviations. To answer these
questions extensions to the normal single species have been developed
[as discussed in @magnusson1995msvpa and references therein
]. These extensions attempt to ecompass
significant processes that affect the species of interest.  A special
type of these approaches are the mimimally realistic models mentioned
above [also referred to as MICE models as discussed
in @plaganyi2014multispecies].

A special modelling framework, Gadget [**G**lobally
  **a**pplicable, area--**d**isaggregated
  **e**cosystem **t**oolbox, described
  in @begley2005gadget], has been developed to aid in the study of
various statistical processes involved in multi--species
assessments. The development of Gadget conceptually dates back to
MULTSPEC [@bogstad1997multspec], which described a multi--species
model for fish and marine mammals in the Barents sea. Many of the
underlying concepts and most of the codebase that forms the Gadget
framework was date back to Bormicon [@stefansson1997bormicon,
  @stefansson1998framework], which was developed as a continuation of a
multi-species sampling program [described in @mrifjolst],
which has been under constant developement and iterations. Notable
iterations include the Fleksibest framework
[@guldbrandsen2002fleksibest], a fork of Bormicon which was
merged back with Bormicon to form Gadget as part of the dst$^2$
project [@dst2].

The dst$^2$ project was multi-objective project funded by the European
Union and began in 1999. The basic
idea was to ``develop fisheries science as far as information will
support'' the development of statistically sound models of marine
ecological processes, ``but strictly no further''. Four main
objectives were the following:

* Collect relevant data and provide objective means to analyse
  them. This included the development of Gadget and a specialised data
  warehouse.
* Validate present stock assessment.
* Evaluate the appropriateness of complicated models when
  providing management advice.
* Evaluate and validate reference points used in management.

A substantial amount of analysis has subsequently been made on Cod in
Icelandic waters using the Gadget modeling framework. Notably
@taylor2007simple which describes a thorough analysis of the
available data and potential weaknesses. In it a single species single
area model is considered. A continuation of this work in a multi-stock
multi-species multi-area setting are described in
@taylor2011gadget. There it is seen that substantial improvements
in the model fit are possible by considering a north -- south division
in the cod stock.

These developments have not been restricted to single--species
models. A number of species have been analysed in a multi--species
setting using Gadget. Notable implementaions include a model of the
Barents sea of cod, capelin, herring and minke whales
@lindstrom2009modelling and subsquently
@howell2010combined where FLR @kell2007flr is used in
connection with Gadget in a harvest control simulation.


```{r, engine="tikz", eval=FALSE}
  \begin{tikzpicture}[auto,
    %decision/.style={diamond, draw=black, thick, fill=white,
    %text width=8em, text badly centered,
    %inner sep=1pt, font=\sffamily\small},
    block_center/.style ={rectangle, draw=black, thick, fill=white,
      text width=8em, text centered,
      minimum height=4em},
    block_left/.style ={rectangle, draw=black, thick, fill=white,
      text width=16em, text ragged, minimum height=4em, inner sep=6pt},
    block_noborder/.style ={rectangle, draw=none, thick, fill=none,
      text width=12em, text centered, minimum height=1em},
    block_assign/.style ={rectangle, draw=black, thick, fill=white,
      text width=12em, text ragged, minimum height=3em, inner sep=6pt},
    block_lost/.style ={rectangle, draw=black, thick, fill=white,
      text width=16em, text ragged, minimum height=3em, inner sep=6pt},
      line/.style ={draw, thick, -latex', shorten >=0pt}]
    % outlining the flowchart using the PGF/TikZ matrix funtion
      \matrix (m) [column sep=1mm,row sep=4mm] {
        \node [block_center] (EDL) {Extract from the data sources}; 
        & 
        & \node [block_center] (init) {Gather information on the
          structure of the ecosystem}; \\      
        \node [block_center] (datafiles) {Prepare data into correct
          format: 
          \begin{itemize}
          \item catch distribution 
          \item stock distribution 
          \item biomass indices
          \item ... 
          \end{itemize}};
        & %\node[block_noborder] (inftype) {Inference method:};  
        &  \node [block_center] (inputfile){Information recorded into input
          files
          \begin{itemize} 
          \item Area 
          \item Time 
          \item Stock(s)
          \item Fleet(s)
          \end{itemize}};\\
        % & \node [inputdim] }\\
        
        \node[block_center] (likelihood) {Define the likelihood
          function and assign weights to likelihood components.};  
        & & \\
         & & \\
        & \node[block_noborder](gadget){Run simulation}; & \\
         & \node[block_noborder](empty){};& \\
        \node[block_noborder](calcLik){Calculate likelihood value}; & & 
        \node[block_noborder](optim){Optimiser};\\  
        & & \\ 
        & 
         \node[block_center](finalmodel){Produce model output};& \\
         \node[block_center](goodness){Goodness of fit}; & &
         \node[block_center](derived){Projections};\\
    };% end matrix
    \begin{scope}[every path/.style=line]
      % paths for enrollemnt rows
      \path (optim) -- (gadget);
      \path (calcLik) -- (optim);
      \path (gadget) -- (calcLik);
      \path (EDL) -- (datafiles);
      \path (datafiles) -- (likelihood);
      \path (likelihood) -- (gadget);
      \path (init) -- (inputfile);
      \path (inputfile) -- (gadget);
      \path (empty) -- (finalmodel);
      \path (finalmodel) -- (goodness);
      \path (finalmodel) -- (derived);
    \end{scope}

 \tikzset{blue dotted/.style={draw=blue!50!white, line width=1pt,
                               dash pattern=on 1pt off 4pt on 6pt off 4pt,
                                inner sep=4mm, rectangle, rounded corners}};
                            \node (first dotted box) [blue dotted, 
                            fit = (gadget) (calcLik) (optim)]
                            {};
                            \node at (first dotted box.center) [
                            inner sep=3mm] {\textbf{Gadget}}; 
  \end{tikzpicture}
  
```
An illustration of a typical stock assessment process and the
types of uncertainty associated with each step within Gadget.

## Description of Gadget
Gadget is essentially three things, an ecosystem simulator, a
likelihood function that takes the output from the ecosystem simulator
and compares to data, and a function mininimizer. Gadget's simulation
module allows for a fairly configurable ecosystem simulation. Its
fundamental unit, a **stock** (or more accurately substock),
represents a group of individuals that is homogenous with respect to
various processes. These processes include growth, predation
(including commercial fisheries) and migration. In this setup
different stages of the life history of a particular species would be
represented as separate stocks and individuals ``moved'' between
stocks when required. The simulation takes place in a set number of
**years** or **observations** and **time-steps** within
an observation. Years can be broken down into a maximum of 12 time-steps. 
The time-steps within the observations allow for the
emulation of the annual cycles of the ecosystem, such as recruitment
and stock migrations.

The stock unit within Gadget is simply a representation of the total
number of individuals within a certain age and length group from one or 
more certain area. When stocks live in multiple areas, they
optionally migrate among areas during definable time-steps. 
In this setup, processes such as fleet
harvest or recruitment can be restricted to take place only in certain
areas. Harvesting of the substocks is defined through fleets that fish
according to effort and (length--based) selection functions.

Gadget's likelihood module processes the output from the ecosystem
simulation after aggregating across certain dimensions. Within the likelihood module,
a number of datasets can be compared to the model output. In addition
to a suite of functions designed to work with different types of
survey indices, various other types of data , can be contrasted to
the model output including compositional data (such as length distributions, age distributions, age- and length-distribution, or proportions mature) or tagging data to name a few. Each dataset is compared to model output at its own aggregation
level, with missing data handled in a robust manner.

In contrast with Gadget, age-based or stock production type stock
assessments require data in a fairly processed form. For instance when
using VPA one requires the total catch in numbers of individuals by
age. However, apart from catches of fin whales in the North Atlantic
@iwcistfin2015, one rarely has all catches by numbers at
age. Therefore the age distribution of catches needs to be
approximated using some combination of age readings, length
distributions, total catches in tonnes and weight at age [as noted
in @hirst2005estimating]. Essentially using VPA requires a two-step
modelling process.

Gadget's function minimizer, based on the negative log--likelihood
(but implemented as a simple sum of squares), varies the
model parameters, runs a full simulation, and calculates a new
output. This process is repeated until a minimum is obtained.  The
model has three alternative optimising algorithms linked to it, a wide
area search *simulated annealing* [@corana1987minimizing],
a local search *Hooke and Jeeves algorithm*
[@HookeJeeves1961] and finally one based on the
Broyden-Fletcher-Goldfarb-Shanno algorithm (*BFGS*).

The simulated annealing and Hooke-Jeeves algorithms are not gradient
based, and there is therefore no requirement for the likelihood
surface to be smooth. Consequently neither of the two algorithms
returns estimates of the Hessian matrix. Simulated annealing is more
robust than Hooke and Jeeves and can find a global optima where there
are multiple optima but needs about 2-3 times the order of magnitude
in iterations than the Hooke and Jeeves algorithm to converge.

BFGS is a quasi-Newton optimisation method that
uses information about the gradient of the function at the current
point to calculate the best direction to look for a better point.
Using this information the BFGS algorithm can iteratively calculate a
better approximation to the inverse Hessian matrix. When compared to
the two other algorithms implemented in Gadget, BFGS is a very local
search compared to simulated annealing and is more computationally
intensive than the Hooke and Jeeves.  However the gradient search in
BFGS is more accurate than the step-wise search of Hooke and Jeeves and
may therefore give a more accurate estimation of the optimum.  The
BFGS algorithm used in Gadget is derived from that presented by
@bertsekas1999nonlinear.


The total objective function to be minimised is a weighted sum of the
different components.  The estimation could be difficult due to groups
of correlated parameters or multiple local optima.  Therefore the
optimisation procedure often involves a combination of the more robust
simulated annealing, to make the results less sensitive to the initial
(starting) values, and the local search algorithms (Hooke and
Jeeves and BFGS), to obtain more accurate final values in the 
neighborhood of the global optimum.


# Bibliography