---
title: "Stock interactions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Rgadget)
```

# Types of interations

As we have already seen there are a number of ways we can define interactions between stocks. Here we will, however, explore further the various different mechanisms implemented in Gadget. These interaction can essentially be grouped into two categories:

* Fish entering stock due to a process in a related stock
  + This can be either movement between stocks or spawning
* Fish is removed from a stock due to predation

We will illustrate these processes using a simple multistock model:

```{r}
gd <- gadget.variant.dir('simple_multistock')

schedule <- 
  expand.grid(year = 1:30, step = 1:4) %>% 
  arrange(year)

gadgetfile('Modelfiles/area',
           file_type = 'area',
           components = list(list(areas = 1,
                                  size = 1,
                                  temperature = schedule %>% 
                                    mutate(area = 1, temperature = 5)))) %>% 
  write.gadget.file(gd)


gadgetfile('Modelfiles/area',
           file_type = 'area',
           components = list(list(areas = 1,
                                  size = 1,
                                  temperature = schedule %>% head(5) %>%  
                                    mutate(area = 1, temperature = 5)))) %>% 
  write.gadget.file(gd)

stock_template <- 
  gadgetstock('stock_1',gd,missingOkay = TRUE) %>% 
  gadget_update('stock',
                livesonareas = 1,
                maxage = 1,
                minage = 1,
                minlength = 0,
                maxlength = 2,
                dl = 1) %>% 
  gadget_update('doesgrow',0) %>% 
  gadget_update('naturalmortality', 0.2) %>% 
  gadget_update('refweight',data=tibble(length=0:2,mean=0:2)) %>% 
  gadget_update('initialconditions',
                normalparam = tibble(age = 1,
                                     area = 1,
                                     age.factor = 1,   
                                     area.factor =1,
                                     mean = 1,
                                     stddev = .1,
                                     alpha = 1,
                                     beta = 1)) %>% 
  gadget_update('doesrenew',
                normalparam = schedule %>% 
                  filter(step == 1) %>% 
                  mutate(area = 1, ## where does the rec. take place
                         age = 1, ## age of recruits
                         number = 1, ## num. of recruits (x10 000)
                         mean = 1, ## mean length
                         stddev = .1, ## std in length
                         alpha = 1, ## a in w = aL^b
                         beta = 1))

```

We will use the stock template to define multiple stocks to illustrate the various possible interactions the user is able to define in a Gadget model. 

# Fish recruits into a stock

## Movement and straying

```{r engine='tikz',echo=FALSE}

\usetikzlibrary{arrows,decorations.pathmorphing,decorations.footprints,fadings,calc,trees,mindmap,shadows,decorations.text,patterns,positioning,shapes,matrix,fit,backgrounds}

\begin{tikzpicture}[inner sep = 7mm]

\path 
%(-8,0) node [shape = circle,draw,fill=black!20] (BG1M) {}
%(-5.5,0) node [shape = circle,draw,fill=black!20] (BG2M) {}
(-3,0) node [shape = circle,draw,fill=black!20] (BG3M) {}
(0.5,0) node [shape = circle,draw,fill=black!50, fill opacity=0.5] (BG1D) {}
(2,0) node [shape = circle,draw,fill=black!50, fill opacity=0.5] (BG2D) {}
(3.5,0) node [shape = circle,draw,fill=black!50, fill opacity=0.5] (BG3D) {}
(-7,5) node [shape = circle,draw,fill=black!50, fill opacity=0.5] (FG1M) {}
(-5.5,5) node [shape = circle,draw,fill=black!50, fill opacity=0.5] (FG2M) {}
(-4,5) node [shape = circle,draw,fill=black!50, fill opacity=0.5] (FG3M) {}
(-.5,5) node [shape = circle,draw,fill=black!50, fill opacity=0.5] (FG1D) {}
(2,5) node [shape = circle,draw,fill=black!50, fill opacity=0.5] (FG2D) {}
(4.5,5) node [shape = circle,draw,fill=black!50, fill opacity=0.5] (FG3D) {};
%(1.5,-1) node [draw,fill=white] (txt1) {Dispersion}
\draw [<->] (BG1M) to (FG1M) ;
\draw [<->] (BG2M) to (FG2M) ;
\draw [<->] (BG3M) to (FG3M) ;
\draw [-] (-1.75,-1) to (-1.75,6.5) ;
\draw [-,dashed] (-10,2.5) to (7,2.5) ;
\draw (-7,6.5) node {EG} ;
\draw (-5.5,6.5) node {WI} ;
\draw (-4,6.5) node {EI} ;
\draw (-.5,6.5) node {EG} ;
\draw (2,6.5) node {WI} ;
\draw (4.5,6.5) node {EI} ;
\draw (-8,-2) node {BG1} ;
\draw (-5.5,-2) node {BG2} ;
\draw (-3,-2) node {BG3} ;
\draw (.5,-2) node {BG1} ;
\draw (2,-2) node {BG2} ;
\draw (3.5,-2) node {BG3} ;
\draw [-,dashed] (-6.25,2.5) to (-6.25,6.5) ;
\draw [-,dashed] (-4.75,2.5) to (-4.75,6.5) ;
\draw [-,dashed] (0.75,2.5) to (0.75,6.5) ;
\draw [-,dashed] (3.25,2.5) to (3.25,6.5) ;
\draw [<->] (BG1D) to (FG1D) ;
\draw [<->] (BG2D) to (FG2D) ;
\draw [<->] (BG3D) to (FG3D) ;
\draw [->] (1,-0.2) to (1.8,-0.2) ;
\draw [<-] (0.7,0.2) to (1.5,0.2) ;
\draw [->] (2.5,-0.2) to (3.25,-0.2) ;
\draw [<-] (2.25,0.2) to (3.0,0.2) ;
%\draw[fill=black!20] (0,-1.25) rectangle (3,-.75);
\draw[style=dashed] (2,0) ellipse (3.5 and 1.5);
%\draw [dashed] (0,0) grid (3,2);
\draw (2,7.5) node {Dispersion} ;
\draw (-5.5,7.5) node {Mixing} ;
%\draw (-10,3.25) node {\makecell{Feeding\\grounds}} ;
%\draw (-10,1.75) node {\makecell{Breeding\\grounds}} ;
\end{tikzpicture}

```


## Maturation

## Spawning

# Consumption 

